---
title: "Mini Data-Analysis Deliverable 1"
output: github_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3.  Make a repository in the <https://github.com/stat545ubc-2023> Organization. You can do this by following the steps found on canvas in the entry called [MDA: Create a repository](https://canvas.ubc.ca/courses/126199/pages/mda-create-a-repository). One completed, your repository should automatically be listed as part of the stat545ubc-2023 Organization.

# Instructions

## For Both Milestones

-   Each milestone has explicit tasks. Tasks that are more challenging will often be allocated more points.

-  Each milestone will be also graded for reproducibility, cleanliness, and coherence of the overall Github submission. 

- While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

- The milestones will be equally weighted. 

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 36 points: 30 for your analysis, and 6 for overall reproducibility, cleanliness, and coherence of the Github submission. 

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset 

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 **(1 point)** Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: cancer_sample 

2: steam_games 

3: flow_sample 

4: parking_meters 

<!----------------------------------------------------------------------------->

1.2 **(6 points)** One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->

```{r}
### EXPLORE **steam_games** HERE ###
# Check the numbers of row and column in steam_games
dim(steam_games)

# Check the names of each column and the first six rows
head(steam_games)

# Check the type of the dataset **(tibble vs. data.frame)**
class(steam_games)

```
---

```{r}
### EXPLORE **cancer_sample** HERE ###
# Check the numbers of row and column in dataset
dim(cancer_sample)

# Check the size of dataset and first few entries of each variable
glimpse(cancer_sample)

# Check the type of the dataset **(tibble vs. data.frame)**
class(cancer_sample)
```
---

```{r}
### EXPLORE **flow_sample** HERE ###
# Check the numbers of row and column in dataset
dim(flow_sample)

# Check the names of each column and the first six rows
head(flow_sample)

# Check the type of the dataset **(tibble vs. data.frame)**
class(flow_sample)
```
---

```{r}
### EXPLORE **parking_meters** HERE ###
# Check the numbers of row and column in dataset
dim(parking_meters)

# Check the size of dataset and first few entries of each variable
glimpse(parking_meters)

# Check the type of the dataset **(tibble vs. data.frame)**
class(parking_meters)
```

<!----------------------------------------------------------------------------->

1.3 **(1 point)** Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 1. What lead you to choose this one? Briefly explain your choice below.

<!-------------------------- Start your work below ---------------------------->
**Dataset selected**: cancer_sample 

**Reason**: My thesis project is also a clinical study. The cancer_sample dataset can help me think about how these variables relate to each other and how the analysis informs about the subjects' condition. 
<!----------------------------------------------------------------------------->

1.4 **(2 points)** Time for a  final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with your dataset. Note it down below.

<!-------------------------- Start your work below ---------------------------->
**Research Question**: Are there significant differences in the mean values of the features between malignant and benign tumors?
<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset 

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 **(12 points)** Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. 

Make sure that you're using dplyr and ggplot2 rather than base R for this task. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

1.Plot the distribution of a numeric variable.
```{r}
# Using the ggplot2 package to visualize the distribution of 'concave_points_mean' in the 'cancer_sample' dataset.

# Start with specifying the data and aesthetic mapping
ggplot(cancer_sample, aes(x= concave_points_mean)) +

  # Create a histogram with 30 bins
  # The bars are colored blue and have an opacity of 80%
  geom_histogram(bins=30, fill = "blue", alpha = 0.8) +

  # Apply a light theme to the background of the plot
  theme_light() +

  # Set the title of the plot
  labs(title = "The Distribution of Mean Concave Points") +

  # Center the title above the plot
  theme(plot.title = element_text(hjust = 0.5))
  
```

4.Explore the relationship between 2 variables in a plot.
```{r}
# Create a scatter plot to explore the relationship of mean radius and mean concave points

# Start by specifying the data and aesthetic mappings for x and y axes
ggplot(cancer_sample, aes(x=radius_mean, y=concave_points_mean)) +

  # Create a scatter plot with data points colored based on diagnosis (either M or B)
  # Set the opacity of the points to 0.6
  geom_point(aes(color=diagnosis), alpha=0.6) + 

  # Apply a minimalistic theme to the plot
  theme_minimal() + 

  # Set the title of the plot and labels for x and y axes
  labs(title="Relationship between Mean Radius and Mean Concave Points",
       x="Radius Mean", y="Concave Points Mean") + 

  # Manually specify the colors for the diagnoses: red for Malignant (M) and blue for Benign (B)
  scale_color_manual(values=c("M"="red", "B"="blue"), name="Diagnosis")

```

6.Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
```{r}
# Plot a boxplot of 'concave_points_mean' based on the diagnosis (malignant vs. benign)

# Using ggplot to start the plot with 'cancer_sample' dataset
ggplot(cancer_sample) +

  # Display a boxplot for 'concave_points_mean' separated by 'diagnosis', with blue fill and 80% opacity
  # Adjust the width of the boxplot to 0.5
  geom_boxplot(aes(x= diagnosis, y = concave_points_mean), width = 0.5, fill="blue", alpha=0.8) +

  # Set plot title and axis labels
  labs(title="Distribution of Concave Points Mean Based on Diagnosis", x="Diagnosis", y="concave_points_mean")+ 

  # Center the title text
  theme(plot.title = element_text(hjust = 0.5))

```

8.Use a density plot to explore any of your variables (that are suitable for this type of plot).
```{r}
# Plotting the distribution of 'radius_mean'

ggplot(cancer_sample) + # select the dataset
  
  # Define the x-axis as radius_mean. We can use a density plot for radius_mean because it is a continuous variable. Set color to blue and opacity to 0.5.
  geom_density(aes(x= radius_mean), fill="red", alpha=0.5) + 
  
  # Create the names of x, y-axes and the whole plot.
  labs(title="Distribution of Radius Mean", x="Radius Mean", y="Density")+ 
  
  # Center the title text
  theme(plot.title = element_text(hjust = 0.5))  

```


2.2 **(4 points)** For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->
* Question 1: Plotting the distribution of the mean of concave points reveals the the pattern and variability of data. Understanding its distribution can provide insights into the nature and severity of the tumors in the dataset. We can use historgram for *concave_points_mean* because it is a continuous variable. Line 184 specifies the dataset and creates the x-axis. In line 188, a histogram with 30 bins is generated, with bars colored blue at 80% opacity. Line 191 applies a light theme to the background of the plot. Line 194 and 197 sets the title of the plot and centers it.  
* Question 4: Plotting the relationship between mean radius and mean concave points explores if a tumor with larger radius has more concave points. After selecting the dataset, line 209 set the x-axis as radius_mean and y-axis as concave_points_mean. Line 213 differentiate the data points by using different colors based on *diagnosis* and set the opacity to 0.6. Minimal theme is applied and titles for the plot, x-axis, and y-axis are set. Line 222 specifies the colors for the diagnoses: red for Malignant (M) and blue for Benign (B).
* Question 6: The boxplot shows the difference in the mean of concave points between malignant and benign tumors. I am curious if the number of concave points on tumor can be a indicator of malignancy. After selecting the dataset in line 234, the code uses **geom_boxplot** function from the **ggplot2** package, with *diagnosis* on the x-axis and *concave_points_mean* on the y-axis. The boxplot is styled with a width of 0.5, a blue fill color, and 80% opacity. The title of the plot, "Distribution of Concave Points Mean Based on Diagnosis", and axis labels are clearly defined, with the title being centered for better visual appeal.
* Question 8: By plotting density plot for mean radius, we can get a smooth and continuous distribution of data for all the patients without the constraints of bins in histogram. I'm curious about where most of mean radius distributes. After selecting the dataset in line 254, the code uses **geom_density** function from the **ggplot2** package, with *diagnosis* on the x-axis. The density is styled with a red fill color, and 50% opacity. The title of the plot, "Distribution of Radius Mean", and axis labels are clearly defined, with the title being centered for better visual appeal.         
<!----------------------------------------------------------------------------->

# Task 3: Choose research questions 

**(4 points)** So far, you have chosen a dataset and gotten familiar with it through exploring the data. You have also brainstormed one research question that interested you (Task 1.4).  Now it's time to pick 4 research questions that you would like to explore in Milestone 2! Write the 4 questions and any additional comments below. 

<!--- *****START HERE***** --->
1. Does the worst radius (radius_worst) of tumors significantly differ between malignant and benign tumors?

> This question explores whether the size of the tumor (in this case, represented by its worst radius) is a potential indicator of its malignancy. If there's a significant difference in the worst radius between the two types of tumors, it might serve as a diagnostic marker.

2. Does the fractal dimension of tumors relate to the concavity in both malignant and benign tumors?

> To explore the relationship between two continous variables, a scatterplot might be a good option. Remember to distinguish the data points by "Diagnosis". I want to explore these two variables because they seem to be related and contribute to malignancy conceptually.

3. Among tumors diagnosed as malignant, which features (e.g., smoothness_mean, compactness_mean, etc.) are most correlated with larger tumor size (e.g., area_mean)?

> If certain features are strongly correlated with larger tumor sizes, it can help prioritize which characteristics to monitor in patients with malignant diagnoses.

4. What is the relationship between mean area (area_mean) and the mean of concave points (concave_points_mean) in both malignant and benign tumors?

> This question investigates if a larger tumor tends to have more concave points and which variable is a more reliable indicator of malignancy.

<!----------------------------->

# Overall reproducibility/Cleanliness/Coherence Checklist 

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. An example of a major continuity error is having a data set listed for Task 3 that is not part of one of the data sets listed in Task 1. 

## Error-free code (3 points) 

For full marks, all code in the document should run without error. 1 point deduction if most code runs without error, and 2 points deduction if more than 50% of the code throws an error. 

## Main README (1 point)

There should be a file named `README.md` at the top level of your repository. Its contents should automatically appear when you visit the repository on GitHub.

Minimum contents of the README file:

-   In a sentence or two, explains what this repository is, so that future-you or someone else stumbling on your repository can be oriented to the repository.
-   In a sentence or two (or more??), briefly explains how to engage with the repository. You can assume the person reading knows the material from STAT 545A. Basically, if a visitor to your repository wants to explore your project, what should they know?

Once you get in the habit of making README files, and seeing more README files in other projects, you'll wonder how you ever got by without them! They are tremendously helpful.

## Output (1 point)

All output is readable, recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

(0.5 point deduction if any of the above criteria are not met. 1 point deduction if most or all of the above criteria are not met.)

Our recommendation: right before submission, delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant. Then, after your final commit and push to Github, CHECK on Github to make sure that everything looks the way you intended!

## Tagged release (0.5 points)

You've tagged a release for Milestone 1. 

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
